---
title: "Breakaway Alpha div Crossfoster"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---


```{r setup, echo=F, cache = FALSE, message=FALSE}
#knitr::opts_chunk$set(dev = c('pdf', 'png'), echo=F, fig.path='../figures/')
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(breakaway))
suppressPackageStartupMessages(library(ape))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(wesanderson))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(ggforce))
suppressPackageStartupMessages(library(beepr))
suppressPackageStartupMessages(library(phyloseq))
```



```{r constants}
```


# Introduction

The data set consist of 53 samples from captive DMRs, 11 cross fostered siblings pairs (22 individuals) and 31 control individuals within current cross-fostered group and/or birth groups of cross-fostered individuals. Here we test if cross-fostered individuals had gut microbiomes that differed in alpha diversity from non-cross-fostered individuals.


```{r read-data, echo=FALSE, message=FALSE, warning=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccic')

metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = "")) 
  ) %>%
  mutate(Asample = paste(AnimalID, sample, sep="_")) %>% # get a new sample ID with letter first
  mutate(Crossfostered = case_when(Origcolony != CurrentCol ~ "Yes",
                                   TRUE ~ "No"))
# get palette
pal <- wes_palette("GrandBudapest2")
pal <- pal[c(2,3)]

#create a pseq object
asvsp <- asvs %>% select(asv, count, sample) %>% 
  left_join(metadata %>% select(sample, Asample)) %>% select(-sample) %>% # gove another sample name starting a letter as phyloseq once gave us problems with numbers
  pivot_wider(names_from = 'Asample', values_from='count', values_fil= 0) %>%
  tibble::column_to_rownames('asv') %>% as.matrix()
metadatap <- metadata %>% 
            mutate_if(is.character,as.factor) %>% tibble::column_to_rownames('Asample')

pseq <- phyloseq(otu_table(asvsp, taxa_are_rows = TRUE), sample_data(metadatap))
rm(asvsp, taxonomyp, metadatap)
```


<p>&nbsp;</p>

# Results


<p>&nbsp;</p>

## Plots

```{r Observed richness, echo=FALSE, fig.cap='**Figure 1. Observed alpha diversity (ASV richness).** Colour by if individual crossfostered or not.', message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
observed_c <- sample_richness(pseq)
observed_c <- summary(observed_c) %>% rename(Asample = sample_names) %>% left_join(metadata %>% select(Asample, Crossfostered, Plate_No))
observed_c %>% 
  ggplot(aes(x = Crossfostered, y = estimate, col = Crossfostered)) +
  geom_violin() +
  geom_sina() + 
  geom_violin() +
  ggforce::geom_sina() +
  ylab("Observed richness") +
  xlab("Cross-fostered") +
  scale_colour_manual( values = pal) +
      theme(axis.text.x = element_text( size =15),
          axis.title = element_text( size =15, face = "bold"),
        panel.background=element_rect(fill = 'white', colour='black'),
        strip.background=element_rect(fill = 'white', colour='white'),
        strip.text = element_text(face = "bold", size=15),
        panel.grid = element_blank(),
        legend.position = "none"
        )  +
  ylim(0, 700)
```

<p>&nbsp;</p>

# Breakaway Richness

```{r breakaway, echo=FALSE, fig.cap='**Figure 2.Estimated alpha diversity (richness) with confidence interval with breakaway, sorted by estimate.** Colour by if individual was crossfostered or not.', message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
# Calculate estimated species richness breakaway
adiv_break <- pseq %>% breakaway()
adiv_break <- summary(adiv_break) %>% rename(Asample = sample_names) %>% left_join(metadata %>% select(Asample, Crossfostered, Plate_No))

adiv_break %>%
  mutate(Asample = fct_reorder(Asample, estimate)) %>%
  ggplot(aes(x = Asample, y = estimate,col = Crossfostered)) +
  geom_point() +
  scale_colour_manual( values = pal) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  coord_flip() +
  xlab('Sample') + 
  ylab('Estimated richness with confidence interval') +
  ggtitle("Breakaway richness")
```


<p>&nbsp;</p>

From `breakaway-vignette`: Don't freak out! Those are wide error bars, but nothing went wrong -- it's just really hard to estimate the true number of unknown species. `breakaway` was developed to deal with this, and to make sure that we account for that uncertainty when we do inference.

<p>&nbsp;</p>


```{r breakawayviolin, echo=FALSE, fig.cap='**Figure 3. Estimated alpha diversity (richness) violins. Colour by if individyual was crossfostered or not.**', message=FALSE, warning=FALSE, fig.height=4, fig.width=5}
adiv_break %>%
  ggplot(aes(x = Crossfostered, y = estimate, col = Crossfostered)) +
  geom_violin() +
  geom_sina() + 
  geom_violin() +
  ggforce::geom_sina() +
  ylab("ASV Richness") +
  xlab("Cross-fostered") +
  scale_colour_manual( values = pal) +
      theme(axis.text = element_text( size =15),
          axis.title = element_text( size =15, face = "bold"),
        panel.background=element_rect(fill = 'white', colour='black'),
        strip.background=element_rect(fill = 'white', colour='white'),
        strip.text = element_text(face = "bold", size=15),
        panel.grid = element_blank(),
        legend.position = "none"
        )  +
  ylim(0, 700)
```
<p>&nbsp;</p>

This is very similar to Observed richness.

<p>&nbsp;</p>

## Hypothesis testing

test the null-hypothesis that samples from the two populations did not differ in richness: fit a model with Population as fixed effect using the betta() function in `breakaway`.

**`r figr('betta', T, type = 'Table')`.** Estimate as a response of being cross-fostered or not.
```{r betta, echo=FALSE}
set.seed(3333)
bt_fixed <- betta(formula = estimate ~ Crossfostered, 
                      ses = error, data = adiv_break)
bt_fixed$table %>% kable()
```

No significant difference.



**`r figr('betta_random', T, type = 'Table')`.** Estimate as a response of being cross-fostered or not including plate number as random factor. 
```{r betta_random, echo=FALSE}
set.seed(5252)
bt_fixed_random <-
  betta_random(formula = estimate ~ Crossfostered | Plate_No, 
               ses = error,  data = adiv_break)
bt_fixed_random$table  %>% kable()
```

Also non-significant.


<p>&nbsp;</p>

# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in `r figr('cite-packages', T, type = 'Table')`.

```{r cite-packages, echo=FALSE, warning=FALSE, cache=FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```


<p>&nbsp;</p>

# References
