---
title: "Model microbiome similarity cross-foster experiment"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---

```{r setup, echo=F, cache = FALSE, message=FALSE}
#knitr::opts_chunk$set(dev = c('pdf', 'png'),echo=F, fig.path='../figures/')
#ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(compositions))
suppressPackageStartupMessages(library(picante))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(sjPlot))
suppressPackageStartupMessages(library(lmerMultiMember))
```

```{r constants}
```


```{r read-data, message=FALSE, warning=FALSE, include=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccic')
metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = "")) 
  )
```


<p>&nbsp;</p>

# Introduction

The data set consist of 53 samples from captive DMRs, cross-fostered individuals (22 individuals) and 31 control individuals within current cross-fostered group and/or birth groups of cross-fostered individuals. Here we model microbiome similarity calculated from Aitchison distance based on CLR-transformed data. 

```{r asvsp, include = FALSE}
asvsp <- asvs %>% #left_join(metadata %>% select(sample, Asample)) %>%
  #select(asv, count, sample) %>%
  pivot_wider(names_from = 'asv', values_from='count', values_fill= 0) %>%
  tibble::column_to_rownames('sample') %>% as.matrix()

# Compute Aitchison distance/similarity on whole ASV table
pseudocount <- 0.65 # for clr-transformation
feature_table_clr <- compositions::clr(asvsp+pseudocount)
aitch_dist <- dist(feature_table_clr, method="euclidean")

# To make the Aitchison distance more comparable to other ecological distances 
# -> convert to a similarity after we re-scale it to be bound between 0 and 1
aitch_sim <- (1/(1+aitch_dist/max(aitch_dist)))

# delte unnessesary objects
rm(aitch_dist,asvsp, feature_table_clr, pseudocount)
```


```{r modeldata, include=FALSE}
# turn into matrix
aitch_sim <- aitch_sim %>% as.matrix()

# replace self-comparisons and upper triangle with NA
aitch_sim[upper.tri(aitch_sim , diag = TRUE)] <- NA

# melt and filter NA
aitch_sim  <- aitch_sim %>% 
  reshape2::melt() %>% filter(!is.na(value)) %>%
  mutate(Var1  = as.character(Var1),
         Var2  = as.character(Var2))

# get group affiliations - same or different
data.dyad <- aitch_sim %>% 
  left_join(metadata %>% select(sample, CurrentCol, Origcolony) %>% 
              rename(Var1 = sample, Origcolony1 = Origcolony, CurrentCol1 = CurrentCol)) %>%
  left_join(metadata %>% select(sample, CurrentCol, Origcolony) %>% 
              rename(Var2 = sample,  Origcolony2 = Origcolony, CurrentCol2 = CurrentCol)) %>%
  mutate(Groupmember = as.factor(case_when(CurrentCol1 == CurrentCol2 ~ 1,
                                 TRUE ~ 0)),
         Related = as.factor(case_when(Origcolony1 == Origcolony2 ~ 1,
                                 TRUE ~ 0)),
         GroupmembBirth = as.factor(case_when(Origcolony1 == CurrentCol2 ~ 1, 
                                              Origcolony2 == CurrentCol1 ~ 1,
                                              TRUE ~ 0))
         ) %>% rename(Microbiome_similarity = value)

# check distribution of data
hist(data.dyad$Microbiome_similarity) # fairly normally dist # proportional values semi-normally distributed limited between 0 and 1
hist(data.dyad$Microbiome_similarity[data.dyad$Groupmember == 1])
hist(data.dyad$Microbiome_similarity[data.dyad$Related  == 1]) 
hist(data.dyad$Microbiome_similarity[data.dyad$GroupmembBirth  == 1])
```

<p>&nbsp;</p>

```{r lmer_model1, include=FALSE}
# calculate m matrix for paired random factors
member_vec <- data.dyad %>% mutate(memb = paste(Var1, Var2, sep = ",")) %>% select(memb) %>% pull()
memb <- lmerMultiMember::weights_from_vector(member_vec)
currgroup_vec <- data.dyad %>% mutate(currgroup = paste(CurrentCol1, CurrentCol2, sep = ",")) %>% select(currgroup) %>% pull()
currgroup <- lmerMultiMember::weights_from_vector(currgroup_vec)
origgroup_vec <- data.dyad %>% mutate(origgroup = paste(Origcolony1, Origcolony2, sep = ",")) %>% select(origgroup) %>% pull()
origgroup <- lmerMultiMember::weights_from_vector(origgroup_vec)

model1 <- lmerMultiMember::lmer(Microbiome_similarity ~ Groupmember * Related +
                                  (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships=list(members=memb, curr=currgroup, orig=origgroup))
```


# Results 

<p>&nbsp;</p>

```{r lmer_model1table, include=FALSE}
# create empty df for estimates
mod1_table <- data.frame(matrix(ncol = 5, nrow =4))
colnames(mod1_table) <- c("Predictors","Estimates" ,"Std.Error", "tvalue", "pvalue")

#
mod1_table$Predictors[4] <- "Groupmember*Related"
mod1_table$Estimates[4] <- coef(summary(model1))[4]
mod1_table$Std.Error[4] <- coef(summary(model1))[4,2]
mod1_table$tvalue[4] <- coef(summary(model1))[4,3]

# model without interaction
model1b <- lmerMultiMember::lmer(Microbiome_similarity ~ Groupmember + Related + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))

coef(summary(model1b)) # estimates fixed factors 

#
mod1_table$Predictors[2] <- "Groupmember"
mod1_table$Estimates[2] <- coef(summary(model1b))[2]
mod1_table$Std.Error[2] <- coef(summary(model1b))[2,2]
mod1_table$tvalue[2] <- coef(summary(model1b))[2,3]

#
mod1_table$Predictors[3] <- "Related"
mod1_table$Estimates[3] <- coef(summary(model1b))[3]
mod1_table$Std.Error[3] <- coef(summary(model1b))[3,2]
mod1_table$tvalue[3] <- coef(summary(model1b))[3,3]

#
mod1_table$Predictors[1] <- "Intercept"
mod1_table$Estimates[1] <- coef(summary(model1b))[1]
mod1_table$Std.Error[1] <- coef(summary(model1b))[1,2]
mod1_table$tvalue[1] <- coef(summary(model1b))[1,3]

# pvalue interaction
mod1_table$pvalue[4] <- anova(model1, model1b)$"Pr(>Chisq)"[2] 

# excluding relatedness
model1c <- lmerMultiMember::lmer(Microbiome_similarity ~ Groupmember  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model1b, model1c) # pval relatness
#pval relatness
mod1_table$pvalue[3] <- anova(model1b, model1c)$"Pr(>Chisq)"[2] 

# excluding groupmember
model1d <- lmerMultiMember::lmer(Microbiome_similarity ~ Related  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model1b, model1d) # pval group member
#pval relatness
mod1_table$pvalue[2] <- anova(model1b, model1d)$"Pr(>Chisq)"[2] 

# write table
#mod1_table %>% 
#  mutate_if(is.numeric, round, digits = 3) %>% 
#  write.table("../lmersimilarity.txt", sep = ",", quote = FALSE, row.names = F)
```



**`r figr('tablemod', T, type = 'Table')`. LMM modelling microbiome similarity (Aitchison) of gut microbiomes of Damaraland mole-rats that were part of the cross-foster experiment.** Shared environment but not relatedness predicts gut microbiome similarity. 
```{r tablemod, echo=FALSE}
mod1_table %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  kable()
```


<p>&nbsp;</p>


```{r model1_plot, fig.cap='**Figure 1. Model estimates of gut microbiome similarity (Aitchison) of cross-fostering experiment Damaraland mole-rats.** Points represent model estimates with 95 % confidence intervals. Microbiome similarity is bound between 0 and 1, with higher values indicating more similar microbial communities. Microbiome similarity are from individuals within the same group (Shared Environment) or different groups (Not Shared Environment), that either have same birth group (Related) or different birth groups (Not Related).', echo=FALSE}
p <- sjPlot::plot_model(model1, type = "int",  ci.lvl = .95)
#summary(model1)

p_dat <- ggplot_build(p)
p_dat <- p_dat[[1]]
p_dat <- p_dat[[2]] %>% select(-PANEL, -flipped_aes,  -xmin, -xmax, -linetype, -width, -alpha, -colour, -fill)
p_dat$Groupmember <- c("Not Shared", "Not Shared", "Shared", "Shared")
p_dat$Related <- c("Not Related", "Related", "Not Related", "Related") 
p_dat$label <- c("1", "2", "3", "4")

p_dat %>% 
  ggplot(aes(x = Related, y = y )) +
  geom_pointrange(aes(x = Related, y = y, ymin = ymin, ymax = ymax), size= 0.5) +
  xlab("Environment") +
  ylab("Microbiome similarity") +
  theme_classic() +
  #ylim(0.55,0.625) +
  facet_grid(~Groupmember, scales = "free", space = "free", switch= "x") + 
  theme(strip.background = element_blank(), 
        panel.grid = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.border = element_blank(), 
        panel.spacing = unit(0, "mm"),
        legend.position='none',
        axis.text = element_text(size=16),
        axis.title = element_text(size=22))
```

<p>&nbsp;</p>


```{r check_model,fig.cap='**Figure 2. Model residual checks.**', echo=FALSE}
# check residuals function
mcheck<-function(obj,...){
  rs<-resid(obj)
  fv<-fitted(obj)
  par(mfrow=c(1,3))
  plot(fv,rs,xlab="FITTED VALUES",ylab="RESIDUALS")
  abline(h=0,lty=2,lwd=2)
  qqnorm(rs,xlab="NORMAL SCORES",ylab="ORDERED RESIDUALS",main="")
  qqline(rs,lty=2,lwd=2)
  hist(resid(obj),xlab="RESIDUALS",ylab="FREQUENCY",main="")
  par(mfrow=c(1,1))
  invisible(NULL)
}

mcheck(model1)
```

It is not a perfect world, but good enough.


<p>&nbsp;</p>



# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in `r figr('cite-packages', T, type = 'Table')`.

```{r cite-packages, echo=FALSE, cache=FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```


<p>&nbsp;</p>

# References

