---
title: "Model phylogenetic microbiome similarity cross-foster experiment"
author: "hanna.bensch@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
   html_document:
        toc: yes
        toc_float:
          collapse: no
        fig_caption: yes
        code_folding: hide
bibliography: 
    - grateful-refs.bib
biblio-style: apalike
link-citations: yes
---

```{r setup, echo=F, cache = FALSE, message=FALSE}
knitr::opts_chunk$set(dev = c('pdf', 'png'),echo=F, fig.path='../figures/')
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, include = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(compositions))
suppressPackageStartupMessages(library(picante))
suppressPackageStartupMessages(library(grateful))
suppressPackageStartupMessages(library(sjPlot))
suppressPackageStartupMessages(library(lmerMultiMember))
suppressPackageStartupMessages(library(phyloseq))
```

```{r constants}
```


```{r read-data, message=FALSE, warning=FALSE, include=FALSE}
asvs <- read_tsv("../data/asv_table.tsv", col_types = 'ccic') # setwd() to "../scripts"
metadata <- read_delim("../data/metadata.csv", delim = ",", col_types = cols(
  .default = col_character(),
  SampleDate = col_date(format = "")) 
  )
spptree <- read.tree("../data/CrossSepptree.tog.tree") 

# make a character of the sample number (had problems with making sample as rows for phyloseq object long time ago)
metadata <- metadata %>% mutate(Asample = paste("A", sample, sep= "_"))
asvs <- asvs %>% left_join(metadata %>% select(sample,  Asample))
```


```{r make-phyloseq object, include=FALSE}
str(asvs)
asvsp <- asvs %>% select(asv, count, Asample) %>% pivot_wider(names_from = 'Asample', values_from='count', values_fill = 0) %>%
  tibble::column_to_rownames('asv') %>% as.matrix()
metadatap <- metadata %>% tibble::column_to_rownames('Asample')
OTU = otu_table(asvsp, taxa_are_rows = TRUE)
samples = sample_data(metadatap)
  
pseq <- phyloseq(OTU, samples) # make pseq object
pseq <- merge_phyloseq(pseq, spptree) # merge with tree
print(pseq)
pseq@phy_tree # tree is rooted

# delte unnessesary objects
rm(metadatap,asvsp, samples,OTU)
```

<p>&nbsp;</p>

# Introduction


The data set consist of 53 samples from captive DMRs, cross-fostered individuals (22 individuals) and 31 control individuals within current cross-fostered group and/or birth groups of cross-fostered individuals.Here we model microbiome similarity of weighted and unweighted UniFrac to compare with out results on analysis on Aitchison distance based on CLR-transformed data.


```{r unifrac, include = FALSE}
# calulcate unifrac
# unweighted
dist_uunifrac <- distance(pseq, "uunifrac")
# weighted
dist_wunifrac <- distance(pseq, "wunifrac")

# -> convert to a similarity
uunifrac_sim <- (1-dist_uunifrac)
wunifrac_sim <- (1-dist_wunifrac)

rm(dist_uunifrac, dist_wunifrac) 

# melt and filter NA
uunifrac_sim  <- uunifrac_sim %>% as.matrix() %>%
  reshape2::melt() %>% filter(!is.na(value)) %>%
  mutate(Var1  = as.character(Var1),
         Var2  = as.character(Var2))
wunifrac_sim  <- wunifrac_sim %>% as.matrix() %>%
  reshape2::melt() %>% filter(!is.na(value)) %>%
  mutate(Var1  = as.character(Var1),
         Var2  = as.character(Var2))

# get group affiliations - same or different
data.dyad <- uunifrac_sim %>% 
  left_join(metadata %>% select(Asample, CurrentCol, Origcolony) %>% 
              rename(Var1 = Asample, Origcolony1 = Origcolony, CurrentCol1 = CurrentCol)) %>%
  left_join(metadata %>% select(Asample, CurrentCol, Origcolony) %>% 
              rename(Var2 = Asample,  Origcolony2 = Origcolony, CurrentCol2 = CurrentCol)) %>%
  mutate(Groupmember = as.factor(case_when(CurrentCol1 == CurrentCol2 ~ 1,
                                 TRUE ~ 0)),
         Related = as.factor(case_when(Origcolony1 == Origcolony2 ~ 1,
                                 TRUE ~ 0)),
         GroupmembBirth = as.factor(case_when(Origcolony1 == CurrentCol2 ~ 1, 
                                              Origcolony2 == CurrentCol1 ~ 1,
                                              TRUE ~ 0))
         ) %>% rename(Microbiome_similarityUU = value) %>%
  left_join(wunifrac_sim %>% rename(Microbiome_similarityWU = value)) %>%
  filter(Var1 != Var2)

# check distribution of data - fairly normally dist - proportional values semi-normally distributed limited between 0 and 1
hist(data.dyad$Microbiome_similarityWU) # 
hist(data.dyad$Microbiome_similarityWU[data.dyad$Groupmember == 1])
hist(data.dyad$Microbiome_similarityWU[data.dyad$Related  == 1]) 
hist(data.dyad$Microbiome_similarityWU[data.dyad$GroupmembBirth  == 1])
hist(data.dyad$Microbiome_similarityUU) 
hist(data.dyad$Microbiome_similarityUU[data.dyad$Groupmember == 1])
hist(data.dyad$Microbiome_similarityUU[data.dyad$Related  == 1]) 
hist(data.dyad$Microbiome_similarityUU[data.dyad$GroupmembBirth  == 1])
```

<p>&nbsp;</p>

```{r lmer_model, include=FALSE}
# calculate m m matrix for paired random factors
member_vec <- data.dyad %>% mutate(memb = paste(Var1, Var2, sep = ",")) %>% select(memb) %>% pull()
memb <- lmerMultiMember::weights_from_vector(member_vec)
currgroup_vec <- data.dyad %>% mutate(currgroup = paste(CurrentCol1, CurrentCol2, sep = ",")) %>% select(currgroup) %>% pull()
currgroup <- lmerMultiMember::weights_from_vector(currgroup_vec)
origgroup_vec <- data.dyad %>% mutate(origgroup = paste(Origcolony1, Origcolony2, sep = ",")) %>% select(origgroup) %>% pull()
origgroup <- lmerMultiMember::weights_from_vector(origgroup_vec)

# weighted
model1 <- lmerMultiMember::lmer(Microbiome_similarityWU ~ Groupmember * Related +
                                  (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships=list(members=memb, curr=currgroup, orig=origgroup))

# un-weighted
model2 <- lmerMultiMember::lmer(Microbiome_similarityUU ~ Groupmember * Related +
                                  (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships=list(members=memb, curr=currgroup, orig=origgroup))
```


# Results 

<p>&nbsp;</p>


## Weighted unifrac
```{r lmer_model1table, include=FALSE}
# create empty df for estimates
mod1_table <- data.frame(matrix(ncol = 5, nrow =4))
colnames(mod1_table) <- c("Predictors","Estimates" ,"Std.Error", "tvalue", "pvalue")

#
mod1_table$Predictors[4] <- "Groupmember*Related"
mod1_table$Estimates[4] <- coef(summary(model1))[4]
mod1_table$Std.Error[4] <- coef(summary(model1))[4,2]
mod1_table$tvalue[4] <- coef(summary(model1))[4,3]

# model without interaction
model1b <- lmerMultiMember::lmer(Microbiome_similarityWU ~ Groupmember + Related + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))

coef(summary(model1b)) # estimates fixed factors 

#
mod1_table$Predictors[2] <- "Groupmember"
mod1_table$Estimates[2] <- coef(summary(model1b))[2]
mod1_table$Std.Error[2] <- coef(summary(model1b))[2,2]
mod1_table$tvalue[2] <- coef(summary(model1b))[2,3]

#
mod1_table$Predictors[3] <- "Related"
mod1_table$Estimates[3] <- coef(summary(model1b))[3]
mod1_table$Std.Error[3] <- coef(summary(model1b))[3,2]
mod1_table$tvalue[3] <- coef(summary(model1b))[3,3]

#
mod1_table$Predictors[1] <- "Intercept"
mod1_table$Estimates[1] <- coef(summary(model1b))[1]
mod1_table$Std.Error[1] <- coef(summary(model1b))[1,2]
mod1_table$tvalue[1] <- coef(summary(model1b))[1,3]

# pvalue interaction
mod1_table$pvalue[4] <- anova(model1, model1b)$"Pr(>Chisq)"[2] 

# excluding relatedness
model1c <- lmerMultiMember::lmer(Microbiome_similarityWU ~ Groupmember  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model1b, model1c) # pval relatness
#pval relatness
mod1_table$pvalue[3] <- anova(model1b, model1c)$"Pr(>Chisq)"[2] 

# excluding groupmember
model1d <- lmerMultiMember::lmer(Microbiome_similarityWU ~ Related  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model1b, model1d) # pval group member
#pval relatness
mod1_table$pvalue[2] <- anova(model1b, model1d)$"Pr(>Chisq)"[2] 

#write table
#mod1_table %>% 
#  mutate_if(is.numeric, round, digits = 3) %>% 
# write.table("../lmersimilarity_wuf.txt", sep = ",", quote = FALSE, row.names = F)
```



**`r figr('tablemod', T, type = 'Table')`. LMM modelling microbiome similarity (weighted unifrac) of gut microbiomes of Damaraland mole-rats that were part of the cross-foster experiment.** Shared environment but not relatedness predicts gut microbiome similarity. 
```{r tablemod1, echo=FALSE}
mod1_table %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  kable()
```


<p>&nbsp;</p>


```{r model1_plotwu, fig.cap='**Figure 1. Model estimates of gut microbiome similarity (weighted unifrac) of cross-fostering experiment Damaraland mole-rats.** Points represent model estimates with 95 % confidence intervals. Microbiome similarity is bound between 0 and 1, with higher values indicating more similar microbial communities. Microbiome similarity are from individuals within the same group (Shared Environment) or different groups (Not Shared Environment), that either have same birth group (Related) or different birth groups (Not Related).', echo=FALSE}
p <- sjPlot::plot_model(model1, type = "int",  ci.lvl = .95)
#summary(model1)

p_dat <- ggplot_build(p)
p_dat <- p_dat[[1]]
p_dat <- p_dat[[2]] %>% select(-PANEL, -flipped_aes,  -xmin, -xmax, -linetype, -width, -alpha, -colour, -fill)
p_dat$Groupmember <- c("Not Shared", "Not Shared", "Shared", "Shared")
p_dat$Related <- c("Not Related", "Related", "Not Related", "Related") 
p_dat$label <- c("1", "2", "3", "4")

pwu <- p_dat %>% 
  ggplot(aes(x = Related, y = y )) +
  geom_pointrange(aes(x = Related, y = y, ymin = ymin, ymax = ymax), size= 0.5) +
  xlab("Environment") +
  ylab(expression(atop("Microbiome similarity", paste("weighted unifrac")))) +
  theme_classic() +
  #ylim(0.55,0.725) +
  facet_grid(~Groupmember, scales = "free", space = "free", switch= "x") + 
  theme(strip.background = element_blank(), 
        panel.grid = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.border = element_blank(), 
        panel.spacing = unit(0, "mm"),
        legend.position='none',
        axis.text = element_text(size=16),
        axis.title = element_text(size=22))
pwu
```

<p>&nbsp;</p>


```{r check_model,fig.cap='**Figure 2. Model residual checks model 1, weighted unifrac.**', echo=FALSE}
# check residuals function
mcheck<-function(obj,...){
  rs<-resid(obj)
  fv<-fitted(obj)
  par(mfrow=c(1,3))
  plot(fv,rs,xlab="FITTED VALUES",ylab="RESIDUALS")
  abline(h=0,lty=2,lwd=2)
  qqnorm(rs,xlab="NORMAL SCORES",ylab="ORDERED RESIDUALS",main="")
  qqline(rs,lty=2,lwd=2)
  hist(resid(obj),xlab="RESIDUALS",ylab="FREQUENCY",main="")
  par(mfrow=c(1,1))
  invisible(NULL)
}

mcheck(model1)
```

Nor not as good as Aitchison.

<p>&nbsp;</p>

## Un-weighted unifrac
```{r lmer_mod2table, include=FALSE}
# create empty df for estimates
mod2_table <- data.frame(matrix(ncol = 5, nrow =4))
colnames(mod2_table) <- c("Predictors","Estimates" ,"Std.Error", "tvalue", "pvalue")

#
mod2_table$Predictors[4] <- "Groupmember*Related"
mod2_table$Estimates[4] <- coef(summary(model2))[4]
mod2_table$Std.Error[4] <- coef(summary(model2))[4,2]
mod2_table$tvalue[4] <- coef(summary(model2))[4,3]

# model without interaction
model2b <- lmerMultiMember::lmer(Microbiome_similarityUU ~ Groupmember + Related + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))

coef(summary(model2b)) # estimates fixed factors 

#
mod2_table$Predictors[2] <- "Groupmember"
mod2_table$Estimates[2] <- coef(summary(model2b))[2]
mod2_table$Std.Error[2] <- coef(summary(model2b))[2,2]
mod2_table$tvalue[2] <- coef(summary(model2b))[2,3]

#
mod2_table$Predictors[3] <- "Related"
mod2_table$Estimates[3] <- coef(summary(model2b))[3]
mod2_table$Std.Error[3] <- coef(summary(model2b))[3,2]
mod2_table$tvalue[3] <- coef(summary(model2b))[3,3]

#
mod2_table$Predictors[1] <- "Intercept"
mod2_table$Estimates[1] <- coef(summary(model2b))[1]
mod2_table$Std.Error[1] <- coef(summary(model2b))[1,2]
mod2_table$tvalue[1] <- coef(summary(model2b))[1,3]

# pvalue interaction
mod2_table$pvalue[4] <- anova(model2, model2b)$"Pr(>Chisq)"[2] 

# excluding relatedness
model2c <- lmerMultiMember::lmer(Microbiome_similarityUU ~ Groupmember  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model1b, model1c) # pval relatness
#pval relatness
mod2_table$pvalue[3] <- anova(model1b, model1c)$"Pr(>Chisq)"[2] 

# excluding groupmember
model2d <- lmerMultiMember::lmer(Microbiome_similarityUU ~ Related  + 
                                   (1|members) + (1|curr) + (1|orig), 
                                data = data.dyad,
                                memberships = list(members=memb, curr=currgroup, orig=origgroup))
anova(model2b, model2d) # pval group member
#pval relatness
mod2_table$pvalue[2] <- anova(model2b, model2d)$"Pr(>Chisq)"[2] 

# write table
mod2_table %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  write.table("../lmersimilarity_uu.txt", sep = ",", quote = FALSE, row.names = F)
```



**`r figr('tablemod2', T, type = 'Table')`. LMM modelling microbiome similarity (un-weighted unifrac) of gut microbiomes of Damaraland mole-rats that were part of the cross-foster experiment.** Shared environment but not relatedness predicts gut microbiome similarity. 
```{r tablemod2, echo=FALSE}
mod2_table %>% 
  mutate_if(is.numeric, round, digits = 3) %>% 
  kable()
```


<p>&nbsp;</p>


```{r model2_plotuu, fig.cap='**Figure 3. Model estimates of gut microbiome similarity (un-weighted unifrac) of cross-fostering experiment Damaraland mole-rats.** Points represent model estimates with 95 % confidence intervals. Microbiome similarity is bound between 0 and 1, with higher values indicating more similar microbial communities. Microbiome similarity are from individuals within the same group (Shared Environment) or different groups (Not Shared Environment), that either have same birth group (Related) or different birth groups (Not Related).', echo=FALSE}
p <- sjPlot::plot_model(model2, type = "int",  ci.lvl = .95)
#summary(model2)

p_dat <- ggplot_build(p)
p_dat <- p_dat[[1]]
p_dat <- p_dat[[2]] %>% select(-PANEL, -flipped_aes,  -xmin, -xmax, -linetype, -width, -alpha, -colour, -fill)
p_dat$Groupmember <- c("Not Shared", "Not Shared", "Shared", "Shared")
p_dat$Related <- c("Not Related", "Related", "Not Related", "Related") 
p_dat$label <- c("1", "2", "3", "4")

puu <- p_dat %>% 
  ggplot(aes(x = Related, y = y )) +
  geom_pointrange(aes(x = Related, y = y, ymin = ymin, ymax = ymax), size= 0.5) +
  xlab("Environment") +
  ylab(expression(atop("Microbiome similarity", paste("un-weighted unifrac")))) +
  theme_classic() +
  #ylim(0.52,0.725) +
  facet_grid(~Groupmember, scales = "free", space = "free", switch= "x") + 
  theme(strip.background = element_blank(), 
        panel.grid = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.border = element_blank(), 
        panel.spacing = unit(0, "mm"),
        legend.position='none',
        axis.text = element_text(size=16),
        axis.title = element_text(size=22))
puu
```

```{r check_model2,fig.cap='**Figure 4. Model residual checks model 2, un-weighted unifrac.**', echo=FALSE}

mcheck(model2)
```

models plots slightly better than weighted

# combine plots for supp
<p>&nbsp;</p>


```{r comb_plot, fig.width= 10, fig.height= 6, fig.cap='**Figure 5. Model estimates of gut microbiome similarity, based on A) weighted unifrac, and B) un-weighted unifrac,  of cross-fostering experiment Damaraland mole-rats.** Points represent model estimates with 95 % confidence intervals. Microbiome similarity is bound between 0 and 1, with higher values indicating more similar microbial communities. Microbiome similarity are from individuals within the same group (Shared Environment) or different groups (Not Shared Environment), that either have same birth group (Related) or different birth groups (Not Related).', echo=FALSE}

ggpubr::ggarrange(pwu +
                    ylab("w-unifrac sim") +
                    xlab("") +
                    theme(axis.text = element_text(size=14),
                          axis.title = element_text(size=14)),
                  puu +
                    ylab("uw-unifrac sim") +
                    theme(axis.text = element_text(size=14),
                          axis.title = element_text(size=14)),
                  labels = c("A", "B"), ncol =1)
```

# R and packages

This analysis was run with `r R.version$version.string`. Versions of packages used can be found in `r figr('cite-packages', T, type = 'Table')`.

```{r cite-packages, echo=FALSE, cache=FALSE}
cite_packages(pkgs ="Session", output = "table") %>%
  kable()
```


<p>&nbsp;</p>

# References

